---
title: "edgeR_KO"
author: "Dennis Amnebrink"
affiliation-title: "Linnaeus University"
title-block-banner: true
date: now
editor: visual
toc: true
prefer-html: true
number-sections: true
format:
  html:
     code-fold: true
     self-contained: true
---

```{r}
SIGNIFICANCE = 0.05
MIN_COUNT = 15
MIN_SAMPLE = 2
```

```{r}

```

```{r}
GEOM_COL_COLOURS = c(
  RColorBrewer::brewer.pal(12, name = 'Paired'))

GEOM_COL_COLOURS1 = c(GEOM_COL_COLOURS, "#D3D3D3")
```

```{r}
#| label: libraries
#| warning: false
library(tidyverse)
library(data.table)
library(edgeR)
library(kableExtra)
library(viridis)
library(patchwork)
library(ggbreak)
library(DT)
```

# Load in data from data2 folder

## Subheading 2

### Subheading 3

```{r}
#| label: file-loading
#| warning: false
#| cache: true
#File to connect NGI ID with our own classification ID, add the µ-values to this file for testing later, or do that in a sep file?
sample_ID <- read_tsv("../data/J.Pinhassi_21_02_sample_info.txt") %>%
  dplyr::select(1:2) %>%
  dplyr::rename(sample = 1, sample_name = 2) %>%
  mutate(sample_name = str_replace(sample_name, "[0-9]*:","")) %>%
  separate(sample_name, c("treatment","timepoint"), sep = ",") %>%
  mutate(treatment = gsub("NT","TN", treatment)) %>%
  mutate(sample_name = paste(timepoint, treatment, sep = "_")) %>%
  separate(treatment, c("treatment","replicate"), sep = -1) %>%
  mutate(tre_rep = gsub("_","", sample_name)) %>%
  mutate(tre_rep = toupper(tre_rep)) %>%
  mutate(day = case_when( timepoint == "t3" ~ 10,
                          TRUE ~ 17
                          )
         )

target_genes <- read_tsv("../data/list_of_genes.tsv") 

#Reading in annotations
eggnogs <- read_tsv("../data2/all_samples.emapper.annotations.tsv", comment = "#")  %>%
  rename(orf = "query") 
  

#Reading in taxonomic annotation of reads
taxonomy <- read_tsv("../data2/out.CAT.contig2classification.names.txt.gz") %>%
  rename(chr = "# contig") %>%
  mutate(superkingdom = gsub(":.*","",superkingdom),
         phylum = gsub(":.*","",phylum),
         class = gsub(":.*","",class),
         order = gsub(":.*","",order),
         family = gsub(":.*","",family),
         genus = gsub(":.*","",genus),
         species = gsub(":.*","",species),
         )

#Count file for all orfs, tpm is calculated per sample
bbmap <- fread("../data2/counts.tsv.gz", sep = "\t") %>%
              filter(count > 0)    %>%                           # removing 0 counts to reduce size of table
              mutate(sample = str_extract(sample, "P[0-9]*_[0-9]*")) #Removing trailing sample information

rm_values <- read_tsv("../data/Rm_values.tsv")

gene_categories <- read_tsv("../data/gene_categories.tsv")

kofamscan <- read_tsv("../data2/prokka.kofamscan.tsv.gz")
 

obj1 <- kofamscan %>%
  mutate(score = as.numeric(score)) %>%
  select(-1) %>% #Removing significance column, not needed
  rename(orf = "gene name") %>%
  group_by(orf) %>% #This grop_by selects best hit of all KO's
  slice_max(score, n = 1) %>%
  ungroup()

obj2 <- kofamscan %>%
  mutate(score = as.numeric(score)) %>%
  select(-1) %>% #Removing significance column, not needed
  rename(orf = "gene name") %>%
  group_by(orf) %>% #This grop_by selects best hit of all KO's
  arrange(desc(score)) %>%
  slice(1) %>%
  ungroup()
```

```{r}
#| cache: true
#| warning: false
#Removing non-prokreads as well as recalculating tpm over prok frac.
bbmap_p <- bbmap %>% 
  semi_join(taxonomy %>% filter(superkingdom %in% c("Bacteria", "Archaea")), 
            by = "chr") %>%
  dplyr::select(-tpm) %>%
   group_by(sample) %>%
            mutate(t = count/length) %>%
            mutate(tpm = t/sum(t)*1e6) %>%
   ungroup()
```

# EdgeR- analysis

```{r}
#| label: edgeR-analysis
#| warning: false
#| cache: true

#| label: edgeR
#| cache: true
#| warning: false
design <- model.matrix(~ 0 + factor(c(1,1,1,2,2,2,3,3,3,4,4,4)))
colnames(design) <- c("warm_control","cold_control","cold_nutrient","warm_nutrient")

#Creating DGElist
dgelist <- bbmap_p %>%
  group_by(orf) %>%
  filter(sum(count) >= MIN_COUNT, n() >= MIN_SAMPLE) %>%
  ungroup() %>%
  semi_join(sample_ID %>% filter(day == '17'), by = 'sample') %>%
  inner_join(eggnogs, by = 'orf') %>%
  group_by(KEGG_ko, sample) %>%
  summarise(count = sum(count), .groups = 'drop') %>%
  ungroup() %>%
  filter(KEGG_ko != "-") %>%
  pivot_wider(names_from = sample, values_from = count, values_fill = 0) %>%
  as.data.frame() %>%
  tibble::column_to_rownames('KEGG_ko') %>%
  DGEList() %>%
  calcNormFactors() %>%
  estimateDisp(design) %>%
  glmQLFit(design, robust = TRUE)

#Creating contrasts
my.contrasts <- makeContrasts(warm_nutrientvscold_control=warm_nutrient-cold_control, levels=design)

qlf.TNvsC <- glmQLFTest(dgelist, contrast=my.contrasts[,"warm_nutrientvscold_control"])
```

```{r}
#| warning: false
#| cache: true
#| 
day17 <- topTags(qlf.TNvsC, n = Inf) %>%
  as.data.frame() %>%
  mutate(day = "17") %>%
  rownames_to_column("KEGG_ko") %>%
  select(-FDR)

edgeR_KO <- day17 %>%
  mutate(fdr = p.adjust(PValue, method = "fdr")) %>%
  mutate(type = 
           case_when(
             (fdr <= SIGNIFICANCE & logFC > 0) ~ "Up",
             (fdr <= SIGNIFICANCE & logFC < 0) ~ "Down",
             TRUE ~"non-DE"
           )) %>%
  mutate(type = forcats::as_factor(type) %>% fct_relevel("Up","non-DE","Down"))

write_tsv(edgeR_KO, "../results/edgeR_KO.tsv")
```

The edgeR analysis revealed 809 genes more abundant in the TN-treatment, and 1006 more abundant in the C-treatment (@tbl-overview)

```{r}
#| label: tbl-overview
#| tbl-cap: "EdgeR summary of genefunctions"
#| warning: false
edgeR_KO <- read_tsv("../results/edgeR_KO.tsv")

kable(edgeR_Preferred_name %>%
  group_by(type) %>%
  tally() %>%
  ungroup() %>%
  rename("No. of genes" = "n"))
```

```{r}
#| label: fig-volcano
#| warning: false
#| eval: false
edgeR_KO %>% 
  ggplot(mapping = aes(x=logCPM, y = logFC, colour=type, size = type)) +
  geom_point()  +
  scale_size_manual(values = c("Down" = 1.5, "Up"=1.5, "non-DE" = 0.3)) +
  scale_colour_manual(values = c("Down" = "Blue","Up" = "Red", "non-DE" = "Black")) +
  theme_minimal() +
  geom_hline(aes(yintercept = DIFF_TRESHOLD), linetype = "dashed", color = "green") +
  geom_hline(aes(yintercept = -DIFF_TRESHOLD), linetype = "dashed", color = "green")
```

Among the target-gene list, only 9 genes were statistically upregulated in the TN-treatment (@tbl-Up). Thus from here we employ a different approach to identify patterns connected to morphology, by looking at large scale patterns and investigating genes of interest in the statistically different gene pools.

```{r}
#| label: tbl-Up
#| tbl-cap: "Target genes among the TN-abundant genes, and their classification."
#| warning: false

#Upload the targent gene lists into kegg and get annotation that way.
#Identifying target genes in the positive DE-expressed genes
targetUp <- edgeR_KO %>%
  filter(type == "Up") 

kable(targetUp %>% arrange(KEGG_ko))
```

```{r}
#| label: tbl-Down
#| tbl-cap: "Target genes among the C-abundant genes, and their classification."
#| warning: false
#| eval: false
#Identifying target genes in the negative DE-expressed genes
targetDown <- edgeR_KO %>%
  filter(type == "Down") 

  kable(targetDown %>% arrange(KEGG_ko))
```

After running and comparing the analysis, it is clear to see that several kegg_KO terms are sharing hits, and thus it becomes ambiguous to retreve information on which to select, due to this Daniel is running a different progfram where this information is provided and a best hit can then be selected.

```{r}
p1 <- bbmap_p %>%
  inner_join(sample_ID %>% filter(day == "17"), by = "sample") %>% # Selecting relevant day
  filter(treatment %in% c("TN","C")) %>% # Selecting relevant treatments
  group_by(orf, treatment) %>%
  summarise(tpm = mean(tpm)) %>% # summarising mean tpm per treatment
  ungroup() %>%
  inner_join(eggnogs %>% distinct(orf, KEGG_ko), by = "orf") %>%
  inner_join(edgeR_KO %>%
    filter(type != "non-DE"), 
    by = "KEGG_ko") %>%
  group_by(KEGG_ko, treatment, type) %>%
  summarise(tpm = sum(tpm)) %>%
  ungroup() %>%
  arrange(desc(tpm)) %>%
  slice(1:100) %>%
  ggplot(mapping = aes(x = KEGG_ko, y = tpm, fill = type)) +
  geom_col() +
  facet_wrap(~ treatment) +
  theme(axis.text.x = element_text(hjust = 1, angle = 90))

ggplotly(p1)
```
