---
title: "PROMAC_target_genes"
author: "Dennis Amnebrink"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output:
  html_document:
      toc: yes
      toc_float:
        collapse: no
      fig_caption: yes
      code_folding: hide
---

The analysis includes the PROMAC mesocosm days 10 and 17. Gene expression for these two timepoints were sequenced and annotated. The list of genes provided by Ashish was then used to search for genes. The genes used can be found at [here](https://github.com/dennisamne/PROMAC_morphology/tree/main/data) at my github page.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(data.table)
library(kableExtra)
```

```{r R_colours, message=FALSE, warning=FALSE}
GEOM_COL_COLOURS = c(
  RColorBrewer::brewer.pal(12, name = 'Paired'))

TREAT_COLOURS <- c("blue","green3","red","yellow4")
```

```{r, variables}
COUNT_FILTER = 5
```

```{r file_loading, cache=TRUE, message=FALSE, warning=FALSE}
#File to connect NGI ID with our own classification ID, add the µ-values to this file for testing later, or do that in a sep file?
sample_ID <- read_tsv("../data/J.Pinhassi_21_02_sample_info.txt") %>%
  dplyr::select(1:2) %>%
  dplyr::rename(sample = 1, sample_name = 2) %>%
  mutate(sample_name = str_replace(sample_name, "[0-9]*:","")) %>%
  separate(sample_name, c("treatment","timepoint"), sep = ",") %>%
  mutate(treatment = gsub("NT","TN", treatment)) %>%
  mutate(sample_name = paste(timepoint, treatment, sep = "_")) %>%
  separate(treatment, c("treatment","replicate"), sep = -1) %>%
  mutate(tre_rep = gsub("_","", sample_name)) %>%
  mutate(tre_rep = toupper(tre_rep)) %>%
  mutate(day = case_when( timepoint == "t3" ~ 10,
                          TRUE ~ 17
                          )
         )
#Reading in annotations
eggnogs <- read_tsv("../data/eggnog_annotations.tsv.gz")

#Reading in taxonomic annotation of reads
taxonomy <- read_tsv("../data/eukulele_phylodb.tsv") %>%
  select(-species) %>%
  rename(species = "genus", genus = "family", family = "order", order = "class", class = "phylum", phylum = "kingdom")  %>%
  mutate_all(function(x) ifelse(str_detect(x, '_X'), NA, x)) %>% #Removing _X-annotations
  mutate_all(function(x) ifelse(str_detect(x, 'environmental'), NA, x)) %>% # Removing environmental - annotations this is considered the same as NA
  mutate_all(function(x) ifelse(str_detect(x, 'Misc'), NA, x)) %>%
  mutate_all(function(x) ifelse(str_detect(x, 'unclassified'), NA, x))
  
  
#Count file for all orfs, tpm is calculated per sample
bbmap <- fread("../data/bbmap_counts.tsv.gz", sep = "\t") %>%
              filter(count > 0)  %>%                               # removing 0 counts to reduce size of table
              mutate(Geneid = str_replace(Geneid, "[0-9]*_",""),   # Joining in chr and geneid to match format of eggnog naming.
                     orf = paste(Chr,Geneid, sep = "_"),
                     sample = str_extract(sample, "P[0-9]*_[0-9]*")) %>% # Removing extra numbers in sample to match sample_ID format
              dplyr::select(-Geneid,-Chr) %>%
              dplyr::select(orf,Start,End,Strand,Length,sample,count,tpm) 

target_genes <- read_tsv("../data/list_of_genes.tsv") 
```

```{r, message=FALSE, warning=FALSE, cache=TRUE}
#Removing non-prokreads as well as recalculating tpm over prok frac.
bbmap_p <- bbmap %>% 
  semi_join(taxonomy %>% filter(domain %in% c("Bacteria", "Archaea")), 
            by = "orf") %>%
  dplyr::select(-tpm) %>%
   group_by(sample) %>%
            mutate(t = count/Length) %>%
            mutate(tpm = t/sum(t)*1e6) %>%
   ungroup()
```

```{r, message=FALSE, warning=FALSE, results=FALSE}
#Figuring out how many genes are matching, and looking at the ones who do not, and why that is.
target_gene_names  <- target_genes %>% distinct(gene) %>% pull(gene)

enog_names <- eggnogs %>% distinct(Preferred_name) %>% pull(Preferred_name)

length(target_gene_names)

target_gene_names[match(target_gene_names, enog_names)]

#Extracting the matching gene names between the two lists
  matches <- enog_names[!is.na(target_gene_names[match(enog_names, target_gene_names)])]
  
#Looking at the names not matching to see if it is errors in naming, 69 matches so only 27 genes missing
  target_genes %>%
    filter(!gene %in% matches) 
  
```
Out of `r length(target_gene_names)` target genes, a total of `r length(matches)` genes were found in the dataset.
<br>
```{r, message=FALSE, warning=FALSE, fig.cap="Target gene expression grouped by larger categories on days 10 and 17."}
bbmap_p %>%
  inner_join(eggnogs, by = "orf") %>%
  inner_join(target_genes %>% rename(Preferred_name = "gene"), by = "Preferred_name") %>%
  inner_join(sample_ID, by = "sample") %>%
  group_by(timepoint, treatment, replicate, sub_category) %>%
  summarise(tpm = sum(tpm)) %>%
  ungroup() %>%
  group_by(timepoint,treatment, sub_category) %>%
  summarise(tpm = mean(tpm), stdev = sd(tpm)) %>%
  ungroup() %>%
  ggplot(mapping = aes(x = sub_category, y = tpm, group = timepoint, fill = timepoint)) +
  geom_col(position = "dodge") +
  geom_errorbar(aes(ymin=tpm - stdev, ymax=tpm + stdev)) + #Why no errorbars??
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  facet_wrap(~ treatment) +
  xlab("categories") +
  ggtitle("Gene categories")
```
 
```{r, count_filtering, message=FALSE, warning=FALSE}
filtering_df <- bbmap_p %>%
  inner_join(eggnogs, by = "orf") %>%
  semi_join(target_genes %>% rename(Preferred_name = "gene"), by = "Preferred_name") %>%
  inner_join(target_genes %>% rename(Preferred_name = "gene"), by = "Preferred_name") %>%
  inner_join(sample_ID, by = "sample") %>%
  group_by(Preferred_name, sample) %>%
  summarise(sum_count = sum(count)) %>%
  ungroup() %>%
  inner_join(sample_ID, by = "sample") %>%
  group_by(timepoint, treatment, Preferred_name) %>%
  summarise(mean_count = mean(sum_count)) %>%
  ungroup() %>%
  group_by(Preferred_name) %>%
  summarise(sum_count = sum(mean_count)) %>%
  ungroup() %>%
  filter(sum_count > (length(unique(sample_ID$treatment)) * length(unique(sample_ID$timepoint)) * COUNT_FILTER)) # This is to require a minimum average of 5 counts per sample
  

filtered_genes <- filtering_df %>% 
  distinct(Preferred_name) %>% 
  pull(Preferred_name)
```
<br>
After applying a filter criteria of a minimum of `r COUNT_FILTER` reads per sample on average, only 
 `r length(filtered_genes)` genes remained.
<br>
```{r, message=FALSE, warning=FALSE, fig.width=30, fig.height=14, fig.cap="Target gene expression."}
# The gene list and genes expressed
bbmap_p %>%
  inner_join(eggnogs, by = "orf") %>%
  inner_join(target_genes %>% rename(Preferred_name = "gene"), by = "Preferred_name") %>%
  inner_join(sample_ID, by = "sample") %>%
  group_by(timepoint, treatment, Preferred_name) %>%
  summarise(tpm = mean(tpm), stdev =sd(tpm)) %>%
  ungroup() %>%
  inner_join(target_genes %>% rename(Preferred_name = "gene")) %>%
  semi_join(filtering_df, by = "Preferred_name") %>%
  ggplot(mapping = aes(x = Preferred_name, y = tpm, group = timepoint, fill = timepoint)) +
  geom_col(position = "dodge") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1, size = 25),
        axis.text.y = element_text(size = 25)) +
  scale_x_discrete(guide = guide_axis(n.dodge=3)) +
  facet_wrap(~ treatment) +
  ggtitle("Target genes")
```
